---
title: "Data"
output: html_document
date: "2026-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The brief

The brief was to:
*You will choose an application domain that your group will work with during Sprint Assessment 0-1.
*Your challenge is to perform a literature review of the resources for helping you do data science in your chosen application domain.

## The team

This is a five person team: *Anna*, *Bobby*, *Jonty*, *Mathilde* and *Max*. We are all experts in R. We have agreed to work in this coding language. We intend to contribute --% to the project.

## Match and Odds Data

We decided to examine English Premier League football data, and through the website "https://footiqo.com/database/leagues/england-premier-league/" we are able to obtain information about the distribution of goals in a match, as well as betting odds set by a bookmaker before the match begins. We currently have data on the 2024/2025 season which is stored on the repository, however for group sprint 1 we will obtain data from 2015 to 2026. We have a mixture of numerical and categorical variables, however we can easily format categorical variables into factors, such as the results of match being "H" for home, "D" for draw, and "A" for away.

The data we have available per match is: 

- Win for home, away, or draw
- Over/under 0.5 goals (0-0 bet)
- O/U 1.5, 2.5, 3.5, 4.5 goals
- Both teams to score Yes/No
- Date

We also add the goal result data as well to use in the classifier, to know the actual outcome of the match after the odds have been set. 

- Full Time Home/Away goal
- 1st Half Home/Away goal
- 2nd Half Home/Away goals
- 1st Half / 2nd Half / Full Time Result

We also added 3 columns ourselves just to make data handling easier: 
- HP and AP are Home Points and Away Points to understand how many points each team received every match
- PRED WIN is what the bookmakers have predicted will happen in the match, either Home win, Draw, or Away win

We aim to use these covariates, along with the FPL player information to create a team "score" that can be used in a classifier to predict the outcomes of matches. 

Using the odds data, we also aim to evaluate how accurately bookmakers can predict the outcomes of matches, and we could create a model and score to describe how likely a given team is to satisfy a bet depending on historical bookmaker data and their accuracy on bets. This creates a classification problem of: "given these odds by a bookmaker and the teams playing, how likely is a bet to come true?". 

We may need recency data, such as position in league, previous successes against the team, previous wins/losses in the season, status of team composition and "performance index" of players. I believe that the available data we have from the site is not enough to create a conclusive model but it can act as a stepping stone. 

It may also be informative to source data with dynamic betting odds over different bookmakers, and see if arbitrage opportunities exist within matches to create hedging and convergence opportunities. However, this requires a lot of data due to the interval of change in odds and may not be readily available or free. 

## Fantasy Premier League Player Data

We found individual player data for each season from 2020/21 to 2025/26 in the repository attached below.

https://github.com/vaastav/Fantasy-Premier-League/tree/master/data

We focused on Fantasy Premier League (FPL) data covering six seasons of the English Premier League. In Fantasy Premier League, users build their own team using real players from the Premier League, with players earning or losing points based on their performances in real matches. For example, players can earn points for scoring goals and lose points for receiving a red card.

The dataset contains information on players from Premier League squads across the six seasons. It includes each player's total FPL points accumulated during each season, along with other information such as minutes played and yellow cards received.

We cleaned the dataset by removing players who recorded 0 minutes played during the season. We also added an additional column identifying which Premier League team each player played for, as this information was not included in the original dataset.

The cleaned dataset can be used to investigate patterns between the FPL points earned by players at a club and that same club's overall performance during the season. In particular, we can investigate whether teams whose players earn more FPL points tend to perform better in the Premier League and finish in a higher league position. As the dataset covers six seasons, these patterns can also be compared across different seasons.

This dataset is suitable for the project because it contains FPL performance information for players across six Premier League seasons. The data includes each player's total FPL points, as well as other information such as minutes played and yellow cards. Having data from multiple seasons allows player performance to be compared across different teams and seasons. The additional team column I added also allows the player data to be grouped by club, which could be useful when investigating the relationship between FPL performance and team performance.

However, FPL points are based on specific individual actions and FPL scoring rules, so they do not directly represent how well a football team or individual player performs overall. Not every aspect of a player's performance is measured when calculating FPL points. For example, a player who covers a lot of distance, makes accurate passes and contributes to their team's ability to win may have a significant impact on the match but receive relatively few FPL points if they do not score, assist, or complete other actions that earn points. Therefore, FPL points may not fully reflect a player's overall contribution to their team. Players can also move between teams during a season, which could make comparisons between players and teams more difficult. Other factors that can affect a team's performance, such as injuries, transfers and changes in management, may also not be represented in the dataset.

The FPL data could be used to investigate whether there is a relationship between the FPL points earned by players at a club and that club's overall Premier League performance. For example, the total FPL points earned by players at each club could be compared with the club's final league position in a particular season. This could help identify whether teams that perform better in the Premier League tend to have players who earn more FPL points.


## Premier League Player Data For seasons 2020/21 to 2024/25

Scraped from fbref using the worldfootballR package. This data contains information on players from Premier League squads across five seasons. It contains realised statistics for each player, detailing every players number of goals, assists, minutes etc. Potentially, we could use this data to engineer features akin to a teams ability rating.

We examined the data from the 2024-25 season and found it to be mostly clean, the only issue was that four players where missing their ages. We manually added this missing data.

## Premier League Match Result Data For seasons 2020/21 to 2024/25

Recovered from `fbref` using the `worldfootballR` package. These csv files contain match results for the English Premier League across five seasons. 

The files are saved to the project's data/ directory and are excluded via .gitignore. To recreate them, run the code in `Jonty Heffaran/Match_result_data.Rmd`. The data is saved in five separate csv files, one for each season.

## Library requirements

This installs missing packages required for this script only. Each script handles it's own requirements, using a simple check-and-install system. 

Note that this does **not** guarantee that the correct version of packages is installed; for that, we have created a `requirements.txt` which contains all of the version numbers of the required packages. This is automated by `packrat`.

We need `fs` to be able to handle directories in a cross-platform way. This makes the whole analysis much more repeatable and is good data science practice.

```{r}
if (!require("worldfootballR")) install.packages("worldfootballR")
library("worldfootballR")
```


## The data

Each member of the team had the task to find different datasets that contained football data. We found many datasets with information about matches in different tournaments and about football players. In the end, we decided to focus on data from the English Premier League, which helped us develop the classification question explained in the `01-report.Rmd` file.

We place raw data in the `data` folder of our root. (Note: we are in the /report directory now). So our file system will look like this:

* /data
  * /data/Cleaned
    * 
    * 
  * /data/untouched
    * 

### Get the data

First create a directory for it to go in:

```{r}
if(!dir.exists(file.path("data"))) dir.create(file.path("data"))

if(!file.exists(file.path("data","Database - Odds - England Premier League - unchanged.csv"))) download.file("https://raw.githubusercontent.com/anna-delpino/Group-sprint-0-Sports-Prediction/refs/heads/main/data/Database%20-%20Odds%20-%20England%20Premier%20League%20-%20LS.csv", destfile=file.path("data","Database - Odds - England Premier League - unchanged.csv"))
oddsdata <- read.csv(file.path("data","Database - Odds - England Premier League - unchanged.csv"), row.names = NULL)

if(!file.exists(file.path("data","dataset - 2020-09-24.csv"))) download.file("https://raw.githubusercontent.com/anna-delpino/Group-sprint-0-Sports-Prediction/refs/heads/main/data/dataset%20-%202020-09-24.csv", destfile=file.path("data","dataset - 2020-09-24.csv"))
dataset2020 <- read.csv(file.path("data","dataset - 2020-09-24.csv"))
```

Download it
TBC

### Get the names for the data

### Process the names

### Data output