---
title: "Data"
output: html_document
date: "2026-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The brief

The brief was to:



## The team

This is a five person team: *Anna*, *Bobby*, *Jonty*, *Mathilde* and *Max*. We are all experts in R. We have agreed to work in this coding language. We intend to contribute --% to the project.

## Library requirements

This installs missing packages required for this script only. Each script handles it's own requirements, using a simple check-and-install system. 

Note that this does **not** guarantee that the correct version of packages is installed; for that, we have created a `requirements.txt` which contains all of the version numbers of the required packages. This is automated by `packrat`.

We need `fs` to be able to handle directories in a cross-platform way. This makes the whole analysis much more repeatable and is good data science practice.

```{r}
if (!require("fs")) install.packages("fs")
library("fs")
```


## The data

Each member of the team had the task to find different datasets that contained football data. We found many datasets with information about matches in different tournaments and about football players. In the end, we decided to focus on data from the English Premier League, which helped us develop the classification question explained in the `01-report.Rmd` file.

We place raw data in the `data` folder of our root. (Note: we are in the /report directory now). So our file system will look like this:

* /data
  * /data/raw
    * 
    * 
  * /data/processed
    * 

### Get the data

First create a directory for it to go in:
```{r}
rawdatadir=path_wd("..","data","raw")
if(!dir.exists(rawdatadir)) dir.create(rawdatadir,recursive = TRUE)
```

Download it
```{r}
download.file("https://github.com/dsbristol/dst/raw/master/data/conn_sample.log",path_wd("..","data","raw","conn_sample.log"))
```

### Get the names for the data

NB We had to scrape this from the website which is not very elegent, but works.

```{r}
cnames=c('ts','uid','id.orig_h','id.orig_p',
    'id.resp_h','id.resp_p','proto','service','duration',
    'orig_bytes','resp_bytes','conn_state','local_orig',
    'missed_bytes','history','orig_pkts','orig_ip_bytes',
    'resp_pkts','resp_ip_bytes','tunnel_parents')
write.table(cnames,
            file=path_wd("..","data","raw","conn_sample.names"),
            row.names=FALSE,
            col.names=FALSE,
            quote=FALSE
            )
```

### Create the processed data output.

```{r}
processeddatadir=path_wd("..","data","processed")
if(!dir.exists(processeddatadir)) dir.create(processeddatadir,recursive = TRUE)

conndata=read.table(path_wd("..","data","raw","conn_sample.log"),
                    as.is=T)
colnames(conndata)=cnames
write.table(conndata,
            path_wd("..","data","processed","conn_sample.tab"),
            quote=FALSE,row.names=FALSE)
```

### Summary:

We have now created our data for further processing, in a simple headed tabular format (which is space separated). This can be easily read into either R or Python.