---
title: "Data"
output: html_document
date: "2026-10-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The brief

The brief was to:



## The team

This is a five person team: *Anna*, *Bobby*, *Jonty*, *Mathilde* and *Max*. We are all experts in R. We have agreed to work in this coding language. We intend to contribute --% to the project.

## Match and Odds Data

We decided to examine English Premier League football data, and through the website "https://footiqo.com/database/leagues/england-premier-league/" we are able to obtain information about the distribution of goals in a match, as well as betting odds set by a bookmaker before the match begins. We currently have data on the 2024/2025 season which is stored on the repository, however for group sprint 1 we will obtain data from 2015 to 2026. We have a mixture of numerical and categorical variables, however we can easily format categorical variables into factors, such as the results of match being "H" for home, "D" for draw, and "A" for away.

The data we have available per match is: 

- Win for home, away, or draw
- Over/under 0.5 goals (0-0 bet)
- O/U 1.5, 2.5, 3.5, 4.5 goals
- Both teams to score Yes/No
- Date

We also add the goal result data as well to use in the classifier, to know the actual outcome of the match after the odds have been set. 

- Full Time Home/Away goal
- 1st Half Home/Away goal
- 2nd Half Home/Away goals
- 1st Half / 2nd Half / Full Time Result

We aim to use these covariates, along with the FPL player information to create a team "score" that can be used in a classifier to predict the outcomes of matches. 

## Library requirements

This installs missing packages required for this script only. Each script handles it's own requirements, using a simple check-and-install system. 

Note that this does **not** guarantee that the correct version of packages is installed; for that, we have created a `requirements.txt` which contains all of the version numbers of the required packages. This is automated by `packrat`.

We need `fs` to be able to handle directories in a cross-platform way. This makes the whole analysis much more repeatable and is good data science practice.

```{r}
if (!require("fs")) install.packages("fs")
library("fs")
```


## The data

Each member of the team had the task to find different datasets that contained football data. We found many datasets with information about matches in different tournaments and about football players. In the end, we decided to focus on data from the English Premier League, which helped us develop the classification question explained in the `01-report.Rmd` file.

We place raw data in the `data` folder of our root. (Note: we are in the /report directory now). So our file system will look like this:

* /data
  * /data/Cleaned
    * 
    * 
  * /data/untouched
    * 

### Get the data

First create a directory for it to go in:
```{r}
dir.create("data", showWarnings = FALSE)
download.file("https://github.com/anna-delpino/Group-sprint-0-Sports-Prediction/blob/main/data/Untouched%20FPL%20Data/Untouched_Fantasy_Premier_League_2025_26_Season_Dataset.csv",destfile = "data/Untouched_Fantasy_Premier_League_2025_26_Season_Dataset.csv",mode = "wb")
if(!dir.exists(rawdatadir)) dir.create(rawdatadir,recursive = TRUE)
```

Download it
TBC

### Get the names for the data

### Process the names

### Data output