---
title: "2021 Player data eda"
author: "J. Heffaran"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(ggplot2)
library(tidyverse)
```


We are going to also have a quick look at the player data from 2020-21 season. First, let us read in our df.

```{r}
df <- read.csv(
"C:/Users/jonty/OneDrive/Documents/GitHub/Group-sprint-0-Sports-Prediction/data/premier_league_player_data_2020_21_to_2024_25/premier_league_2020-21_player_data.csv")
```

Now let us have a look at the headers, and column names.

```{r}
print(head(df))
#print(names(df))
```
I have provided a cribsheet with explanation of the column names below, note this arduous work was generated by a LLM, and is only for reference.

 1. Player / season information

| Column | Meaning |
|---|---|
| `Season_End_Year` | The year the season ended. E.g. `2025` means the 2024–25 season. |
| `Squad` | The player's team. |
| `Comp` | The competition. |
| `Player` | Player's name. |
| `Nation` | Player's nationality. |
| `Pos` | Player's position. |
| `Age` | Player's age during the season. |
| `Born` | Player's birth year. |
---

 2. Playing time

| Column | Meaning |
|---|---|
| `MP_Playing` | Matches played. |
| `Starts_Playing` | Matches started. |
| `Min_Playing` | Total minutes played. |
| `Mins_Per_90_Playing` | Minutes played expressed as 90-minute games. |
---

 3. Goals and assists

| Column | Meaning |
|---|---|
| `Gls` | Goals scored. |
| `Ast` | Assists. |
| `G+A` | Goals + assists. |
| `G_minus_PK` | Goals excluding penalties. |
| `PK` | Penalties scored. |
| `PKatt` | Penalties attempted. |
---

 4. Discipline

| Column | Meaning |
|---|---|
| `CrdY` | Yellow cards. |
| `CrdR` | Red cards. |
---

 5. Ball progression

| Column | Meaning |
|---|---|
| `PrgC_Progression` | Progressive carries: carrying the ball forwards significantly. |
| `PrgP_Progression` | Progressive passes: passes that significantly advance the ball. |
| `PrgR_Progression` | Progressive passes received: receiving the ball in positions that significantly advance play. |
---

 6. Per-90 statistics

| Column | Meaning |
|---|---|
| `Gls_Per` | Goals per 90 minutes. |
| `Ast_Per` | Assists per 90 minutes. |
| `G+A_Per` | Goals + assists per 90 minutes. |
| `G_minus_PK_Per` | Non-penalty goals per 90 minutes. |
| `G+A_minus_PK_Per` | Non-penalty goals + assists per 90 minutes. |
---

 7. Expected goals (xG) and expected assists (xAG)

| Column | Meaning |
|---|---|
| `xG_Expected` | Expected goals. |
| `npxG_Expected` | Expected goals excluding penalties. |
| `xAG_Expected` | Expected assists. |
| `npxG+xAG_Expected` | Non-penalty expected goals + expected assists. |
---

 8. Expected statistics per 90

| Column | Meaning |
|---|---|
| `xG_Per` | Expected goals per 90 minutes. |
| `xAG_Per` | Expected assists per 90 minutes. |
| `xG+xAG_Per` | Expected goals + expected assists per 90 minutes. |
| `npxG_Per` | Non-penalty expected goals per 90 minutes. |
| `npxG+xAG_Per` | Non-penalty expected goals + expected assists per 90 minutes. |

Now we will explore a quick summary of the data:

```{r}
summary(df)
```
Note: in previous analysis, we added the birth year and age for four players.

I'm interested in goals scored since trying to model the number of goals in a game could be a method we use to model the result of matchs.

The goals data seems to be skewed. We will plot a histogram of the data to see if this is the case.

```{r}
library(ggplot2)
ggplot(df, aes(x = Gls)) +
geom_histogram(binwidth = 1) + labs(
title = "Distribution of Goals Scored",
x = "Goals Scored",
y = "Number of Players"
) 
```


We see that many players in our dataset never score, and a few score a lot of goals. Lets now use some classic football measures: a 10 goal a season player is a term used to describe a usually quite average, if not reliable striker. We will use bin sizes now to have a look at the number of such players.

```{r}
df %>%mutate(
    Goals_Bin = cut(
      Gls,
      breaks = c(-1, 0, 7, 12, 19, Inf),
      labels = c("0", "1–7", "8–12", "13–19", "20+" )
    )
  ) %>%
  ggplot(aes(x = Goals_Bin)) + 
  geom_bar() +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    vjust = -0.5
  ) +
  geom_bar() +
  labs(
    title = "Distribution of Goals Scored in 24-25",
    x = "Goals",
    y = "Number of Players"
  ) +
  theme_minimal()
```


We see there a quite a few 10 goal a season players. There are also few exceptionally high scoring players, only two players scoring 20+ goals this season.

Let us see how the number of goals scored is distributed by position. We would expect the more attacking the position, the more goals are scored.

```{r}
ggplot(df, aes(x = Pos, y = Gls)) +
  geom_boxplot() +
  labs(
    title = "Goals by Position",
    x = "Position",
    y = "Goals"
  ) +
  theme_minimal()
```


Here we see that Forwards (FW) definitely have the greatest mean amount of goals scored. Note that some players play in multiple positions. A particularly weird combination is forwards whom also play in defense. I expect this will be wide players who switch between playing as wingers and wing backs depending on the system being employed on a certain match day.

Let us check this, we will filter the dataset to show the 10 players with the highest amounts of minutes whose positions are denoted FW,DF:

```{r}
top10 <- df %>%
filter(Pos == "FW,DF" ) %>%
arrange(desc(Min_Playing)) %>%
slice_head(n = 10)

top10
```


Okay, my original conjecture involving wingers was incorrect. We instead have present only two such players, whom I know are both forwards most of the time, and who have scored two goals between them, and so have a position mean of 1.

In the data, we have both the number of goals a player scores, as well as the expected number of goals that they will score. We want to see how well these are correlated:

First we will perform a scatter plot of these two variables.

```{r}
ggplot(df, aes(x = xG_Expected, y = Gls)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Goals Scored vs Expected Goals",
    x = "Expected Goals (xG)",
    y = "Goals Scored") +
  theme_minimal()
```


This appears to be a pretty linear relationship, higher xG tends to be correlated with higher amount of goals scored. We will now test this correlation coefficient:

```{r}
print(cor(df$Gls, df$xG_Expected))
```
A coefficient of 0.932 indicates very highly correlated data.

