---
title: "Assessment Document"
output: html_document
date: "2026-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Import Data

Possible data can be found here: https://football-data.co.uk/data.php
England football data: https://football-data.co.uk/englandm.php

Here we can find Excel files containing the necessary data. 

Note: the Excel file downloaded from the website is .csv so we need to save it as an .xsl or .xsls file onto our computer to import it into R

Looking at the Premier League England 2025/2026 season football data I can see that the first row represents the labels of the data. Some examples of labels are: Home/Away team, Referee, Division, Time, Date, etc.

This means that the data has already been labelled for us???

To do: Upload the data into R.

# Option 1
Steps: 
Environment > Import dataset > From Excel 

```{r}
library(readxl) #package that imports Excel files to RStudio
E0 <- read_excel("C:/Users/lenovo/OneDrive - University of Bristol/2026_2027/DST/Group-sprint-0-Sports-Prediction/Anna Del Pino/E0.xlsx") #change this if we move the data set into another folder
```
#Option 2

Create a connection path from the file to R. This is usually used for large datasets.


## Read the data

```{r}
head(E0)
ncol(E0)
```
Where I found the data I was able to also see the descriptions of the features from the following link: https://football-data.co.uk/notes.txt

This tells us what the features mean. Using `ncol()` we can see the number of these. Since there are so many variables I won't write them out but you can find their meanings by following the link.



## Clean the data
We need to put this data into a matrix/vector so that we can analyse it.
Since we know that the first row are the labels, we could create a vector of labels.

```{r}
letsee<- labels(E0)
print(letsee)
```
As we saw from before, there are a lot of features in this dataset which can be shown form the list above.

Now we have a vector of labels which represent our vector of parameters.
To fit a model with this data we also need to separate the data into a matrix.
Note: the dataset is a combination of qualitative and quantitative data which a matrix can only contain only one type of data. Hence we can't actually use a matrix but we will have look.

```{r}
#we need to make the categorical covariates as factors
E0$Date <- as.factor(E0$Date)
E0$Time <- as.factor(E0$Time)
E0$HomeTeam <- as.factor(E0$HomeTeam)
E0$AwayTeam <- as.factor(E0$AwayTeam)
E0$FTR <- as.factor(E0$FTR)
covariates<-as.matrix(E0)
```
The issue here is that the matrix is made up of a combination of numeric and categorical covariates. Hence to sort this out we will have to change the categorical covariates into a factor. 
Idea:
Identify all qualitative entries > Classify them w.r.t. their label (beta) (eg. Arsenal <-> Home team) > assign each qualitative entry to a digit (eg. Arsenal <-> 1, Man. United <-> 2, etc.) > Apply this to the dataset i.e. put tem back in their places > Done.

```{r}

```

## Conclusion

We decided that my data provided isn't as useful as my other teammates since it doesn't contain as many datapoints and the features found in my dataset are different to what we need for our prediction question of which premier league team will win the tournament i.e. which team will end the tournament with the most points.
