---
title: "Assessment Document"
output: html_document
date: "2026-09-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Import Data

Possible data can be found here: https://football-data.co.uk/data.php
England football data: https://football-data.co.uk/englandm.php

Here we can find Excel files containing the necessary data. 

Note: the Excel file downloaded from the website is .csv so we need to save it as an .xsl or .xsls file onto our computer to import it into R

Looking at the Premier League England 2025/2026 season football data I can see that the first row represents the labels of the data. Some examples of labels are: Home/Away team, Referee, Division, Time, Date, etc.

This means that the data has already been labelled for us???

To do: Upload the data into R.

# Option 1
Steps: 
Environment > Import dataset > From Excel 

```{r}
library(readxl) #package that imports Excel files to RStudio
E0 <- read_excel("C:/Users/lenovo/OneDrive - University of Bristol/2026_2027/DST/Group-sprint-0-Sports-Prediction/Anna Del Pino/E0.xlsx") #change this if we move the data set into another folder
```
#Option 2

Create a connection path from the file to R. This is usually used for large datasets.


## Read the data

```{r}
head(E0)
ncol(E0)
```
Where I found the data I was able to also see the descriptions of the features from the following link: https://football-data.co.uk/notes.txt

This tells us what the features mean. Using `ncol()` we can see the number of these. Since there are so many variables I won't write them out but you can find their meanings by following the link.



## Clean the data
We need to put this data into a matrix/vector so that we can analyse it.
Since we know that the first row are the labels, we could create a vector of labels.

```{r}
letsee<- labels(E0)
print(letsee)
```
As we saw from before, there are a lot of features in this dataset which can be shown form the list above.

Now we have a vector of labels which represent our vector of parameters.
To fit a model with this data we also need to separate the data into a matrix.
Note: the dataset is a combination of qualitative and quantitative data which a matrix can only contain only one type of data. Hence we can't actually use a matrix but we will have look.

```{r}
covariates<-as.matrix(E0)
print(covariates)
```
The issue here is that the matrix is made up of a combination of numeric and categorical covariates. Hence to sort this out we will have to change the categorical covariates into a factor. 
Idea:
Identify all qualitative entries > Classify them w.r.t. their label (beta) (eg. Arsenal <-> Home team) > assign each qualitative entry to a digit (eg. Arsenal <-> 1, Man. United <-> 2, etc.) > Apply this to the dataset i.e. put tem back in their places > Done.

```{r}

```

## Analysis of the data

EXAMPLE 
```{r}
summary(E0)
```

This is showing us the usual `summary()` statistical values for each of the labels (as if they were different RVs). This means that we could plot each label to see how they interact with eachother.


```{r}
plot(letsee, covariates )
```
I will try to use the method 
